Design and implementaiton of a live coding environment

### 2017-10-01 live/paper.tex
Evaluating previews section

@@

@@

In this section, we present a formalisation of a live coding environment for a small,
expression-based programming language that supports \kvd{let} binding, member invocations
and $\lambda$ abstractions. This is the necessary minimum for data exploration as described
in the previous section and it excludes constructs such as a mechanism for defining new objects.
%
\begin{equation*}
\begin{array}{lcl}
e &=& \kvd{let}~x = e~\kvd{in}~e ~|~ \lambda x\rightarrow e ~|~ e.m(e, \ldots, e) ~|~ x ~|~ n
\end{array}
\end{equation*}
%
Here, $m$ ranges over member names, $x$ over variables and $n$ over primitive values such as 
numbers. Function values can be passed as arguments to methods (provided by a type provider), but 
for the purpose of this paper, we do not need to be able to invoke them directly.

@@

\paragraph{Elements of the graph.} The nodes of the graph represent individual operations
to be computed. In our design, the nodes themselves are used as keys, so we attach a unique 
\emph{symbol} to some of the nodes. That way, we can create two unique nodes representing, 
for example, access to a member named \ident{take} which differ in their dependencies.

@@

\begin{figure*}[t]
\vspace{-0.5em}
\begin{equation*}
\begin{array}{ll}
\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = &(1)~~\\ 
\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta}(e_i)\\
\quad \textnormal{and}~\bndclr{v} = \Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\\
\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\}
\\[0.75em]
\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = &(2)\\ 
\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta}(e_i)\\
\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\hspace{-200em}.\\
\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
\\[0.75em]
\ident{bind}_{\Gamma, \Delta}(\kvd{let}~x=e_1~\kvd{in}~e_2) = \bndclr{v}, (\{\bndclr{v}\} \cup V \cup V_1, E \cup E_1)&(3)\\
\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind}_{\Gamma,\Delta}(e_1)\\
\quad \textnormal{let}~\Gamma_1 = \Gamma \cup \{(x,\bndclr{v_1})\} \\
\quad \textnormal{let}~\bndclr{v}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e_2)
\end{array}
\begin{array}{ll}
\ident{bind}_{\Gamma, \Delta}(n) = \bnd{val}(n), (\{ \bnd{val}(n) \}, \emptyset) }&(4)
\\[0.75em]
\ident{bind}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)\quad \textnormal{when}~\bndclr{v} = \Gamma(x)&(6)
\\[1em]
\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(7)\\
\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}\\
\quad \textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e)\\
\quad \textnormal{and}~\bndclr{v} = \Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\\
\quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body}) 
\\[1em]
\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(8)\\
\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}\\
\quad \textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e)\\
\quad \textnormal{and}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\notin \ident{dom}(\Delta)\\
\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(s, x), s~\textnormal{fresh}\\
\quad \textnormal{let}~e = (\bndclr{v},\bndclr{v_0},\blbl{body}) 
\\[0.5em]
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{Rules of the binding process, whic constructs a dependency graph for an expression.}
\label{fig:binding-rules}
\end{figure*}

@@

\begin{figure}[!b]
\begin{equation*}
\begin{array}{l}
\Delta_{i}(\bknd{mem}(m), [(\bndclr{v_0},\blbl{arg}(0)),\ldots, (\bndclr{v_n},\blbl{arg}(n))]) = \bndclr{v}\\
\quad \textnormal{for all}~\bnd{mem}(m, s) \in V\\
\quad \textnormal{such that}~(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E~\textnormal{for}~i\in 0,..,n
\\[0.75em]
\Delta_{i}(\bknd{fun}(x), [(\bndclr{v_1},\blbl{body})]) = \bndclr{v}\\
\quad \textnormal{for all}~\bnd{fun}(x, s) \in V\\
\quad \textnormal{such that}~(\bnd{fun}(x, s), \bndclr{v_1}, \blbl{body}) \in E
\\[0.75em]
\Delta_{i}(\bkndclr{v}) = \Delta_{i-1}(\bkndclr{v})\quad \textnormal{(otherwise)}
\end{array}
\end{equation*}
\caption{Updating the node cache after binding a new graph}
\label{fig:loop}
\end{figure}

@@

\section{Evaluating previews}
\newpage

@@

>>>>

% ==================================================================================================
% TODO: Change values from 'n' to 'o' to indicate they are objects and can have members
% ==================================================================================================

@@

\newpage
~

@@

In this section, we present a formalisation of a live coding environment for a small,
expression-based programming language that supports \kvd{let} binding, member invocations
and $\lambda$ abstractions. This is the necessary minimum for data exploration as described
in the previous section.

It excludes constructs such as a mechanism for defining new objects as we assume that those
are imported from the context through a mechanims such as type providers.
%
\begin{equation*}
\begin{array}{lcl}
e &=& \kvd{let}~x = e~\kvd{in}~e ~|~ \lambda x\rightarrow e ~|~ e.m(e, \ldots, e) ~|~ x ~|~ n
\end{array}
\end{equation*}
%
Here, $m$ ranges over member names, $x$ over variables and $n$ over primitive values such as 
numbers. Function values can be passed as arguments to methods (provided by a type provider), but 
for the purpose of this paper, we do not need to be able to invoke them directly.

@@

\begin{figure*}[t]
\vspace{-0.5em}
\begin{equation*}
\begin{array}{ll}
\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = &(1)~~\\ 
\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta}(e_i)\\
\quad \textnormal{and}~\bndclr{v} = \Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\\
\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\}
\\[0.75em]
\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = &(2)\\ 
\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta}(e_i)\\
\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\hspace{-200em}.\\
\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
\\[0.75em]
\ident{bind}_{\Gamma, \Delta}(\kvd{let}~x=e_1~\kvd{in}~e_2) = \bndclr{v}, (\{\bndclr{v}\} \cup V \cup V_1, E \cup E_1)&(3)\\
\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind}_{\Gamma,\Delta}(e_1)\\
\quad \textnormal{let}~\Gamma_1 = \Gamma \cup \{(x,\bndclr{v_1})\} \\
\quad \textnormal{let}~\bndclr{v}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e_2)
\end{array}
\begin{array}{ll}
\ident{bind}_{\Gamma, \Delta}(n) = \bnd{val}(n), (\{ \bnd{val}(n) \}, \emptyset) }&(4)
\\[0.75em]
\ident{bind}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)\quad \textnormal{when}~\bndclr{v} = \Gamma(x)&(6)
\\[1em]
\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(7)\\
\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}\\
\quad \textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e)\\
\quad \textnormal{and}~\bndclr{v} = \Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\\
\quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body}) 
\\[1em]
\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(8)\\
\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}\\
\quad \textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e)\\
\quad \textnormal{and}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\notin \ident{dom}(\Delta)\\
\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(s, x), s~\textnormal{fresh}\\
\quad \textnormal{let}~e = (\bndclr{v},\bndclr{v_0},\blbl{body}) 
\\[0.5em]
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{Rules of the binding process, which constructs a dependency graph for an expression.}
\label{fig:binding-rules}
\end{figure*}

% --------------------------------------------------------------------------------------------------
\paragraph{Elements of the graph.} The nodes of the graph represent individual operations
to be computed. In our design, the nodes themselves are used as keys, so we attach a unique 
\emph{symbol} to some of the nodes. That way, we can create two unique nodes representing, 
for example, access to a member named \ident{take} which differ in their dependencies.

@@

@@

\begin{figure}[!b]
\vspace{-0.5em}
\begin{equation*}
\begin{array}{l}
\Delta_{i}(\bknd{mem}(m), [(\bndclr{v_0},\blbl{arg}(0)),\ldots, (\bndclr{v_n},\blbl{arg}(n))]) = \bndclr{v}\\
\quad \textnormal{for all}~\bnd{mem}(m, s) \in V\\
\quad \textnormal{such that}~(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E~\textnormal{for}~i\in 0,..,n
\\[0.75em]
\Delta_{i}(\bknd{fun}(x), [(\bndclr{v_1},\blbl{body})]) = \bndclr{v}\\
\quad \textnormal{for all}~\bnd{fun}(x, s) \in V\\
\quad \textnormal{such that}~(\bnd{fun}(x, s), \bndclr{v_1}, \blbl{body}) \in E
\\[0.75em]
\Delta_{i}(\bkndclr{v}) = \Delta_{i-1}(\bkndclr{v})\quad \textnormal{(otherwise)}
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{Updating the node cache after binding a new graph}
\label{fig:loop}
\end{figure}

\begin{figure*}
\begin{equation*}
\qquad\qquad\quad\begin{array}{l}
\inference[(lift-expr)~]
  {\bndclr{v} \Downarrow \llbracket e \rrbracket_\Gamma}
  {\bndclr{v} \Downarrow_{\ident{lift}} \llbracket e \rrbracket_\Gamma}
\\
\\
\inference[(lift-prev)~]
  {\bndclr{v} \Downarrow p}
  {\bndclr{v} \Downarrow_{\ident{lift}} \llbracket p \rrbracket_\emptyset}
\\
\\
\inference[(val)~]
  {~}
  {\bnd{val}(n) \Downarrow n }
\\
\\
\inference[(var)~]
  {~}
  {\bnd{var}(x) \Downarrow \llbracket x \rrbracket_x}
  \\[0.5em]~
\end{array}
\qquad
\begin{array}{l}
\inference[(fun-val)~]
  { (\bnd{fun}(x, s), \bndclr{v}, \blbl{body}) \in E & \bndclr{v} \Downarrow p }
  { \bnd{fun}(x, s) \Downarrow \lambda x\rightarrow p }
\\
\\
\inference[(fun-bind)~]
  { (\bnd{fun}(x, s), \bndclr{v}, \blbl{body}) \in E & \bndclr{v} \Downarrow \llbracket e \rrbracket_{x} }
  { \bnd{fun}(x, s) \Downarrow \lambda x\rightarrow e }
\\
\\
\inference[(fun-expr)~]
  { (\bnd{fun}(x, s), \bndclr{v}, \blbl{body}) \in E & \bndclr{v} \Downarrow \llbracket e \rrbracket_{x,\Gamma} }
  { \bnd{fun}(x, s) \Downarrow \llbracket \lambda x\rightarrow e \rrbracket_\Gamma }
\\[0.5em]~
\end{array}
\end{equation*}
\begin{equation*}  
\qquad\qquad\begin{array}{l}
\inference[(mem-val)~]
  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E & \bndclr{v_i} \Downarrow p_i 
    & p_0.m(p_1, \ldots, p_k) \rightsquigarrow p }
  { \bnd{mem}(m, s) \Downarrow p }
\\  
\\
\inference[(mem-expr)~]
  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E & \exists j\in\{0\ldots k\}.\bndclr{v_j}\diagup\hspace{-1em}\Downarrow p_j
   & \bndclr{v_i} \Downarrow_{\ident{lift}} \llbracket e_i \rrbracket_{\Gamma_i}  }
  { \bnd{mem}(m, s) \Downarrow \llbracket e_0.m(e_1, \ldots, e_k) \rrbracket_{\Gamma_0,\ldots,\Gamma k} }
\\
\end{array}
\end{equation*}

\caption{Rules that define evaluation of previews over a dependency graph for an expression}  
\label{fig:eval}
\end{figure*}

@@

\section{Evaluating previews}
\label{sec:previews}

The mechanism for constructing dependency graphs defined in Section~\ref{sec:formal} makes it
possible to provide live previews when editing code without recomputing the whole program 
each time the source code changes.

The nodes in the dependency graph correspond to individual operations that will be performed
when running the program. When the dependencies of an operation do not change while editing 
code, the subsequent dependency graph will reuse a node used to represent the operation.

Our live editor keeps a map from graph nodes to live previews, so a new preview only needs to be
computed when a new node appears in the dependency graph (and the user moves the cursor to a 
code location that corresponds to the node). This section describes how previews are evaluated.

\paragraph{Previews and delayed previews.}
As discussed in Section~\ref{sec:formal}, the body of a function cannot be easily evaluated to
a value if it uses the bound variable. We do not attempt to ``guess'' possible arguments and,
instead, provide a full preview only for sub-expressions with free variables bound by a let binding.
For a function body that uses the bound variable, we obtain a \emph{delayed preview}, which is
an expression annotated with a list of variables that need to be provided before the expression
can be evaluated. We use the following notation:
%
\begin{equation*}
\begin{array}{rcll}
p&\hspace{-0.25em}\in\hspace{-0.25em}&n~|~\lambda x\rightarrow e&\quad(\textit{Fully evaluated previews})\\
d&\hspace{-0.25em}\in\hspace{-0.25em}&p~|~\llbracket e \rrbracket_\Gamma&\quad(\textit{Evaluated and delayed previews})\\
\end{array}
\end{equation*}
%
Here, $p$ ranges over fully evaluated values. It can be either a primitive value (such as number,
string or an object) or a function value with no free variables. A possibly delayed preview $d$ 
can then be either evaluated preview $p$ or an expression $e$ that requires variables $\Gamma$.
For simplicity, we use an untyped language and so $\Gamma$ is a list of variables $x_1, \ldots, x_n$.

\paragraph{Evaluation and splicing.}
In this paper, we omit the specifics of the underlying programming langauge and we focus on the
live coding mechanism. However, we assume that the language is equipped with an evaluation 
reduction $e \rightsquigarrow p$ that reduces a closed expression $e$ into a value $p$.

For delayed previews, we construct a delayed expression using splicing. For example, assuming
we have a delayed previews $\llbracket e_0 \rrbracket_x$ and $\llbracket e_1 \rrbracket_y$. 
If we need to invoke a member $m$ on $e_0$ using $e_1$ as an argument, we construct a new 
delayed preview $\llbracket e_0.m(e_1) \rrbracket_{x, y}$. This operation is akin to expression
splicing from meta-programming [X,Y] and can be more formally captured by Contextual Modal Type 
Theory (CMTT) as outlined below.

\paragraph{Evaluation of previews.}
The evaluation of previews is defined in Figure~\ref{fig:eval}. Given a dependency graph $(V, E)$,
we define a relation $\bndclr{v}\Downarrow d$ that evaluates a sub-expression corresponding to 
the node $\bndclr{v}$ to a (possibly delayed) preview $d$.

The auxiliary relation $\bndclr{v}\Downarrow_{\ident{lift}} d$ always evaluates
to a delayed preview. If the ordinary evaluation returns a delayed preview, so does the auxiliary
relation (\emph{lift-expr}). If the ordinary evaluation returns a value, the value is wrapped
into a delayed preview requiring no variables (\emph{lift-prev}).

Graph node representing a value is evaluated to a value (\emph{val}) and a graph node representing
an unbound variable is reduced to a delayed preview that requires the variable and returns its
value (\emph{var}).

For member access, we distinguish two cases. If all arguments evaluate to values (\emph{member-val}),
then we use the evaluation relation $\rightsquigarrow$, immediately evaluate the member access and 
produce a value. If one of the arguments is delayed (\emph{member-expr}), because the member access 
is in the body of a lambda function, then we produce a delayed member access experession that
requires the union of the variables required by the individual arguments.

Finally, the evaluation of function values is similar, but requires three cases. If the body can 
be reduced to a value with no unbound variables (\emph{fun-val}), we return a lambda function that
returns the value. If the body requires only the bound variable (\emph{fun-bind}), we return a 
lambda function with the delayed preview as the body. If the body requires further variables,
the result is a delayed preview containing a function.

\paragraph{Caching previews.}
For simplicity, the relation $\Downarrow$ in Figure~\ref{fig:eval} does not specify how previews 
are cached and linked to graph nodes. In practice, this is done by maintaining a lookup table
$\bndclr{v} \rightarrow p$ from graph nodes $\bdnclr{v}$ to (possibly delayed) previews $p$.
Whenever the $\Downarrow$ relation is used to obtain a preview for a graph node, we first 
attempt to find an already evaluated preview using the lookup table. If the preview has not
been previously evaluated, we evaluate it and add it to the lookup table.

This means that evaluated previews can be reused in two ways. First, multiple nodes can 
depend on one sub-graph in a single dependency graph (e.g.~if the same sub-expression appears
twice in the program). Second, the keys of the lookup table are graph nodes and nodes are 
reused when a new dependency graph is constructed after the user edits the source code.

\paragraph{Semantics of delayed previews.}
The focus of this paper is on the design and implementation of a live coding environment, 
but it is worth noting that the structure of delayed previews is closely linked to the work
on Contextual Modal Type Theory (CMTT) [X] and comonads [Y].

In CMTT, $\lbrack \Psi \rbrack A$ denotes that a proposition $A$ is valid in context $\Psi$,
which is closely related to our delayed preivews written as $\llbracket A \rrbracket_\Psi$.
CMTT defines rules for composing context-dependent propositions that would allow us to express
the splicing operation used in (\emph{mem-expr}). In categorical terms, the context-dependent
proposition can be modelled as an indexed comonad [X]. The evaluation of a preview with no 
context dependencies (built implicitly into our evaluation rules) corresponds to the counit
operation of a comonad and would be explicitly written as $\llbracket A \rrbracket_\emptyset \rightarrow A$.

@@

Something about how dependency graph bits are reused correctly, i.e. if there is a source code
change that can change the result, the graph will change.

Something about various code edit operations that will not cause recomputation.
